Ecosystem size and complexity dictate riverine biodiversity

### 2021-03-16 manuscript_sciad.Rmd
edits: maintext, minor

Here, we show that ecosystem size and complexity play comparable roles in controlling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Materials and Methods** and **Supplementary Materials**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Materials and Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale [@comteFishDispersalFlowing2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats; see **Materials and Methods** for watershed definition). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific robust regression models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes region-specific slopes of ecosystem size and complexity (**Materials and Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Materials and Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions. The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) at a log-log scale (i.e., power-law), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity obeys the general scaling law along the two orthogonal dimensions of ecosystem structure: size and complexity. In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Materials**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across geographic regions with distinct fish fauna suggest that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

Using theory and extensive community data, our study provides the strong evidence that ecosystem size and complexity play equally important roles in regulating riverine biodiversity. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in scaling biodiversity patterns [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This research was supported by National Science Foundation through a Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through a Water Quality Benefits grant (EPA-G2015-STAR-A1), and the University of North Carolina at Greensboro through a start-up fund.

>>>>

Here, we show that ecosystem size and complexity play comparable roles in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

The proposed theory provided important insights into how ecological communities are structured in branching networks; however, empirically testing the predictions is extremely difficult because it requires metacommunity-level replications. To confront this logistical challenge, we compiled fish community data across two geographic regions: Hokkaido Island in Japan and the Midwestern US (Midwest). These regions are located in comparable latitude ranges (**Figure 1D**) but support distinct fish communities (**Tables S5 and S6**). Therefore, this data set provides an excellent opportunity to examine the generality of our theoretical predictions. After careful data selection (see **Materials and Methods** and **Supplementary Materials**), we estimated $\alpha$, $\beta$, and $\gamma$ diversity (asymptotic species richness; **Materials and Methods**) at `r nrow(wsd_subset[[1]]) + nrow(wsd_subset[[2]])` watersheds (`r nrow(wsd_subset[[1]])` in Hokkaido and `r nrow(wsd_subset[[2]])` in the Midwest), each of which comprised $\ge$ 10 sites of presence-absence fish community data (a total of `r nrow(point_subset[[1]]) + nrow(point_subset[[2]])` sites). These watersheds are small enough (`r round(min(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` to `r round(max(c(wsd_subset[[1]]$area, wsd_subset[[2]]$area)))` km^2^) to assume that fishes can disperse therein at a multi-generation time scale [@comteFishDispersalFlowing2018], while posing challenges to traverse across watersheds (the ocean or lentic habitats; see **Materials and Methods** for watershed definition). We combined this data set with geospatial information, including watershed characteristics (watershed area and branching probability), climates (annual mean temperature and annual cumulative precipitation), and land use patterns (the fraction of agricultural land use and dam density). Using this data set, we assessed whether observed effects of the macro-scale factors are consistent across the two geographic regions by developing global and region-specific robust regression models. In the global model, we assumed that effects of ecosystem size (watershed area) and complexity (branching probability) are constant across the two regions (i.e., fixed slopes). Meanwhile, the region-specific model assumes different slopes of ecosystem size and complexity between the regions (**Materials and Methods**). We compared the support of these competing models using the Bayes factor, a measure of the strength of evidence in favor of one model over the alternative. In our definition (**Materials and Methods**), the Bayes factor of $> 1$ indicates the support for the global model over the region-specific model.

We found decisive support for the global model in explaining $\gamma$ diversity (Bayes factor = `r round(fit$gamma$bayes_factor, 1)`) despite the substantial difference in fish fauna between the study regions (**Tables S5 and S6**). The estimated $\gamma$ diversity linearly increased with increasing watershed area (ecosystem size) and branching probability (ecosystem complexity) on a log-log scale (i.e., power-law), and these patterns were consistent across the study regions (**Table 1 and Figure 4**). Thus, watershed-scale riverine biodiversity likely obeys the general scaling law along the two orthogonal dimensions of ecosystem structure, i,e, size and complexity. In particular, the effect of ecosystem complexity was striking in its magnitude. Average predictive comparisons [@gelmanAveragePredictiveComparisons2007] revealed that an expected increase of $\gamma$ diversity per 0.1 branching probability was `r op(ap_est$sim_estimate[2] * 0.1)` $\pm$ `r op(ap_est$sim_se[2] * 0.1)` species (**Materials and Methods** and **Supplementary Materials**); this level of increase in $\gamma$ diversity requires `r round(ap_est$sim_estimate[2] * 0.1 / ap_est$sim_estimate[1])` km^2^ in the watershed area (see **Table S7** for the estimated average predictive comparisons). This result is reasonable because numerous studies have shown that tributaries are the primary source of environmental heterogeneity in river networks [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018] - the sole theoretical assumption required for the dual scaling of $\gamma$ diversity to emerge. Further, the consistent patterns across the geographic regions with distinct evolutionary histories imply that ecological traits of species in metacommunities may have a minor effect on the scaling relationships, as predicted by our theory.

@@

Using theory and extensive community data, our study provides the strong evidence that ecosystem size and complexity play equally important roles in shaping riverine biodiversity. In particular, the emerging complexity-diversity relationship points to several important avenues for riverine biodiversity conservation. First and foremost, there is now a clear need to explicitly consider the dimension of ecosystem complexity to achieve successful conservation. Human-induced habitat alterations, including flow regulation [@poffHomogenizationRegionalRiver2007], habitat fragmentation [@grillMappingWorldFreeflowing2019], and stream burial [@elmoreDisappearingHeadwatersPatterns2008], may compromise or restrict access to the diverse habitats that complex branching networks may support. Hence, it is imperative to recognize the role of branching complexity and minimize the homogenizing effects of human activities. Second, the complexity perspective may provide insights into the spatial planning of conservation efforts. Riverine reserves and local restoration are increasingly recognized as an effective management tool, and designing a spatial network of conservation sites in rivers is an area of active research [@koningNetworkGrassrootsReserves2020]. While large reserves or restoration sites are undoubtedly important, our results indicate the potential of coordinated networks of local conservation sites in protecting biodiversity. For example, synergies of multiple small reserves may emerge at the watershed scale when ecologically-distinct tributaries are involved in the design, as evidenced by the recent successful conservation of tropical fishes in Thailand's Salween basin [@koningNetworkGrassrootsReserves2020].

While the prevailing evidence supports the importance of ecosystem size in biodiversity relationships [@hubbellUnifiedNeutralTheory2001; @triantisIslandSpeciesareaRelationship2012; @iwasakiEvaluatingRelationshipBasinscale2012; @saboRoleDischargeVariation2010], ecosystem complexity has not received the attention it deserves. However, the ubiquity of scale-invariant complexity across terrestrial [@turnerLandscapeEcologyTheory2015] and aquatic ecosystems [@rodriguez-iturbeFractalRiverBasins2001] calls for more research embracing the two orthogonal dimensions of the ecosystem's geometric structure. Our fundamental findings should apply beyond branching ecosystems because ecosystem size and complexity both represent a physical 'template' providing a wide spectrum of niche opportunities for living organisms. Further, the proposed framework may be extended to other aspects of biodiversity; for example, a logical next step is to explore how these ecosystem properties regulate functional diversity [@petcheyFunctionalDiversityFD2002; @suHumanImpactsGlobal2021] and shifts in community assembly processes [@chaseStochasticCommunityAssembly2010]. Recognizing the dual control of biodiversity broadens viable options for spatial planning of protected areas or restoration sites, thereby helping conserve biodiversity from societal demands that threaten it. The generalization of our findings in spatially complex ecosystems represents a frontier for future research.

@@

We are grateful to numerous people who have been involved in the collection of fish community data. We thank Michio Fukushima and Hirokazu Urabe for providing information on the Hokkaido Freshwater Fish Database HFish and the fish monitoring data at protected watersheds in Hokkaido. We thank Jacques Finlay for his constructive comments on the earlier version of the manuscript. This material is based upon work supported by the National Science Foundation through a Water Sustainability and Climate Program Observatory grant (EAR‐1209402): REACH (Resilience under Accelerated Change) and through the Division of Environmental Biology (DEB 2015634), Environmental Protection Agency through a Water Quality Benefits grant (EPA-G2015-STAR-A1), and the University of North Carolina at Greensboro through a start-up fund.
